---
title: "p8105_hw3_ab6168"
author: "Angelica Bailey"
date: "2025-10-06"
output: github_document
editor_options: 
  chunk_output_type: inline
---

## Problem 1

```{r, message=FALSE}
library(tidyverse)
library(p8105.datasets)
library(patchwork)
```

Loading instacart dataset

```{r}
data("instacart")
```
<br>

**Description**:

The `instacart` data set contains information on grocery orders placed in 2017 by 131,209 unique users and includes 1,384,617 observations. There are a total of 15 variables describing the details of the order, product, and customer behavior. Each row in the data set is a product from an order. Key variables include:

-   `order_id` – Unique identifier for each order

-   `user_id` - Unique customer identifier

-   `product_id` and `product_name` - Identify the specific item purchased

-   `add_to_cart_order` - The sequence in which each product was added to the cart

-   `reordered` - Indicates whether product had been purchased by the user before

-   `order_dow` and `order_hour_of_day` - Represent when the order was placed

-   `aisle` and `department` - Describe the product's category

A few observations:

-   A total of `r instacart |> summarise(n_distinct(product_name))` unique products have been ordered.

-   From the table below we can see the three most ordered products are bananas, strawberries, and spinach.

```{r}
knitr::kable(instacart |> count(product_name, sort = TRUE) |> 
               slice_head(n = 10))
```
<br>
<br>

**How many aisles are there, and which aisles are the most items ordered from?**

```{r}
#getting number of unique aisles
instacart |> summarise(n_distinct(aisle))
```

```{r}
#finding aisles with the most items ordered
instacart |> count(aisle, sort = TRUE)
```

There are **134 aisles**. Most items are ordered from the **Fresh Vegetables** and **Fresh Fruits** aisle.

<br>

**Plot showing the number of items ordered in each aisle.**

```{r}
#making bar chart
instacart |> 
  count(aisle, sort = TRUE) |>  #count of items for each aisle
  filter(n > 10000) |> 
  ggplot(aes(x = aisle, y = n)) +
  geom_col(fill = "darkgreen") +
  coord_flip() +  # flip for readability
  labs(
    title = "Number of Items Ordered per Aisle (Over 10,000 Orders)",
       x = "Aisle",
       y = "Number of Items Ordered"
    )
```

<br>

**Three most popular items in each of the aisles “baking ingredients”, “dog food care”, and “packaged vegetables fruits”:**

```{r}
knitr::kable(
  instacart |>  
  filter(aisle == c("baking ingredients", 
                      "dog food care", 
                      "packaged vegetables fruits")
         ) |> 
  count(aisle, product_name, sort = TRUE) |>  
  group_by(aisle) |>  
  slice_max(order_by = n, n = 3)
)

```

<br>

**Mean hour of the day at which Pink Lady Apples and Coffee Ice Cream are ordered on each day of the week:**

```{r}
knitr::kable(instacart |> 
  filter(
    product_name %in% c("Pink Lady Apples", "Coffee Ice Cream")) |>  
  group_by(product_name, order_dow) |>  
  summarise(
    mean_hour = round(mean(order_hour_of_day),1), .groups = "drop") |> 
  pivot_wider(        
    names_from = order_dow, 
    values_from = mean_hour)
)
```
<br>
<br>

## Problem 2

Tidying Zillow datasets

```{r, message=FALSE}
zori_df =
  read_csv("data/zori_NYC.csv") |> 
  select(-StateName, -State, -City, -Metro, - RegionType) |> 
  #renaming variables
  rename(
    region_id = "RegionID",
    size_rank = "SizeRank",
    zip_code = "RegionName",
    county = "CountyName"
    ) |> 
  mutate(
    county = str_remove(county," County")
  ) |>
  pivot_longer(
    cols = "2015-01-31":"2024-08-31",
    names_to = "date",
    values_to = "rent_price"
  )

zip_df = 
  read_csv("data/Zip Codes.csv") |> 
  janitor::clean_names() |> 
  select(-state_fips, -file_date)

#joining datasets
join_zillow_df = 
  left_join(zori_df, zip_df, by = c("zip_code", "county")) |> 
  drop_na(rent_price)
```
<br>

**How many ZIP codes are observed 116 times? How many are observed fewer than 10 times? Why are some ZIP codes are observed rarely and others observed in each month?**

```{r}
#How many ZIP codes are observed 116 times?

join_zillow_df |> 
  count(zip_code) |> 
  filter(n == 116)
```

There are **48 zip codes** that are observed 116 times.

```{r}
#How many ZIP codes are observed fewer than 10 times?

join_zillow_df |> 
  count(zip_code) |> 
  filter(n < 10)
```

There are **26 zip codes** that are observed fewer than 10 times.

Some ZIP codes may be observed rarely because there is a lack of information on the rent prices for those zip codes depending on the area. There may be little to no rent properties in certain areas.

<br>

**Table showing average rental price by borough and year**

```{r}
avg_rent = join_zillow_df |> 
  #separating date
  separate(date,
    into = c("year", "month","day"), sep = "-") |> 
  group_by(county, year) |> 
  summarize(average_rent = round(mean(rent_price),1), .groups = "drop") |> 
  pivot_wider(
    names_from = county,
    values_from = average_rent
  )

knitr::kable(avg_rent)
```

First, we can see that Richmond county does not have data from 2015 to 2019. The most significant trend is the steep increase in rents across most boroughs from 2021 to 2024. From 2019 to 2020, we see that prices in boroughs like Manhattan, Brooklyn, and Queens dipped before increasing back in 2021, reflecting the market disruption around 2020.

<br>

**Make a plot showing NYC Rental Prices within ZIP codes for all available years. Your plot should facilitate comparisons across boroughs.**

```{r, message = FALSE}
#average rent per zip per year
zip_avg = join_zillow_df |> 
  separate(date,
           into = c("year", "month","day"), sep = "-") |> 
  group_by(county, zip_code, year) |> 
  summarize(mean_rent = mean(rent_price, na.rm = TRUE), .groups = "drop")

#creating plot to compare rent across boroughs
plot1 = ggplot(zip_avg, aes(x = year, y = mean_rent, group = zip_code, color = county)) +
  geom_line(alpha = 0.4) +
  theme(legend.position = "none") +
  geom_smooth(aes(group = county, color = county), method = lm, se = FALSE, size = 1.0) +
  facet_wrap(~county, scales = "free_y") +
  scale_x_discrete(breaks = c("2015", "2024")) +
  labs(
    title = "NYC Rental Prices by ZIP Code and Borough",
    subtitle = "Each line represents a ZIP code; smoothed trends show borough averages",
    x = "Year",
    y = "Average Rent",
    color = "Borough"
  )

plot1
    
```

Some significant elements of this plot are that in Manhattan and Brooklyn, the spread of the rent prices across zip codes is large. In Manhattan, some ZIP codes have average rents approaching \$8,000, while others are closer to \$4,000. On the other hand, the lines for the Bronx, Queens, and Richmond are more close around their average trend lines. While there is still variation, the difference between the most and least expensice neighborhoods are significantly smaller.

<br>

**Compute the average rental price within each ZIP code over each month in 2023. Make a reader-friendly plot showing the distribution of ZIP-code-level rental prices across boroughs**

```{r}
#computing average rent within each zip in 2023
rent_2023 = join_zillow_df |> 
  separate(date,
           into = c("year", "month","day"), sep = "-") |> 
  filter(year == 2023) |> 
  group_by(county, zip_code, month) |> 
  summarize(avg_monthly_rent = mean(rent_price, na.rm = TRUE), .groups = "drop") |> 
  mutate(
    month = case_when(
      month == "01" ~ "Jan",
      month == "02" ~ "Feb",
      month == "03" ~ "Mar",
      month == "04" ~ "Apr",
      month == "05" ~ "May",
      month == "06" ~ "Jun",
      month == "07" ~ "Jul",
      month == "08" ~ "Aug",
      month == "09" ~ "Sep",
      month == "10" ~ "Oct",
      month == "11" ~ "Nov",
      month == "12" ~ "Dec"
    )
  ) |> 
  mutate(month = factor(month, levels = month.abb, ordered = TRUE))

#creating plot showing zip code average rent for each month in 2023
plot2 <- ggplot(rent_2023, aes(x = month, y = avg_monthly_rent, 
                      group = zip_code, color = county )) +
  geom_line(alpha = 0.5) +
  theme(legend.position = "none") +
  facet_wrap(~ county, scales = "free_y") +
  scale_color_brewer(palette = "Set1") +
  scale_x_discrete(breaks = c("Jan", "Dec")) +
  labs(
    title = "ZIP-Code-Level Average Rents by Month in 2023",
    x = "Month",
    y = "Average Rent",
    color = "Borough"
  )

plot2
  
```

From this plot showing average NYC rental prices each month in 2023 across boroughs, we can see that Kings and New York county have more steady prices for each zip code throughout the year, however, for Richmond and Bronx county, there is more fluctuation in prices.

<br>

**Combining Plots**

```{r}
combined_plot <- plot1/plot2
combined_plot
```

Saving plot

```{r, message = FALSE}
ggsave(
  filename = "combined_plot.png",  
  plot = combined_plot,            
  path = "results"
)
```
<br>
<br>

## Problem 3

Cleaning demographic dataset

```{r, message=FALSE}
demo_df = 
  read_csv("data/nhanes_covar.csv") |> 
  filter(row_number() > 3) 
colnames(demo_df) <- demo_df[1, ] #assigning first row as column names
demo_df <- demo_df[-1, ] #removing first row
demo_df =
  demo_df |> janitor::clean_names() |> 
  filter(age >= 21) |> 
  drop_na(sex, age, bmi, education) |> 
  mutate(sex = as.character(sex), 
         sex = case_match(
           sex,
           "1" ~ "male",
           "2" ~ "female"),
         education = as.factor(education),
         seqn = as.numeric(seqn)
  )
```
<br>

Cleaning accelerometer dataset

```{r}
accel_df =
  read.csv("data/nhanes_accel.csv") |> 
  janitor::clean_names() |> 
  pivot_longer(
    cols = min1:min1440,
    names_to = "minute",
    values_to = "mims"
  ) |> 
  mutate(minute = gsub("min", "", minute))
```

Merging datasets

```{r}
mims_df =
  inner_join(demo_df, accel_df)
```
<br>

**Produce a reader-friendly table for the number of men and women in each education category:**

```{r}
sex_educ =
  demo_df |> 
  count(sex, education) |> 
  pivot_wider(names_from = sex, 
              values_from = n) |> 
  mutate(education = as.numeric(as.character(education)),
         education =
           case_match(
             education,
             1 ~ "Less than High School",
             2 ~ "High School Equivalent",
             3 ~ "More than High School"
           ))

knitr::kable(sex_educ, title = "Number of Participants by Education Level and Sex")
```

There are more males than females who have a high school equivalent education. The number of females who've done less than high school is similar to the number of males who've done less than highschool. This is the same case for those who've completed more than highschool.

<br>

**Create a visualization of the age distributions for men and women in each education category:**

```{r}
#setting up data frame for plot
demo_plot_df =
  demo_df |> 
  mutate(age = as.numeric(age),
         education = as.numeric(as.character(education)),
         education =
           case_match(
             education,
             1 ~ "Less than High School",
             2 ~ "High School Equivalent",
             3 ~ "More than High School"
           ))

#plotting histogram
ggplot(demo_plot_df, aes(x = age, fill = sex)) +
  geom_histogram(binwidth = 10, position = "dodge") +
  # Create a separate plot for each education category
  facet_wrap(~ education) +
  labs(
    title = "Age Distribution by Education and Gender",
    x = "Age",
    y = "Count of Participants"
  )
  

```
<br>
Some key observations from this plot are that the 'More than High School' category holds the largest number of participants and skewed towards a younger demographic. There is a prominent peak for females in the 20-30 age bracket. In the other two groups, the age distribution is more evenly spread out across all age ranges. There is no single dominant age group. The overall number of participants in each age group is generally low.

<br>
<br>

**Looking at total acitivity over the day**

Aggregating mims across minutes for each participant:
```{r}
activity_df = mims_df |> 
  group_by(seqn, sex, age, bmi, education) |> 
  summarise(
    total_activity = sum(mims),
    .groups = "drop"
  ) |> 
  mutate(education = as.numeric(as.character(education)),
         education =
           case_match(
             education,
             1 ~ "Less than High School",
             2 ~ "High School Equivalent",
             3 ~ "More than High School"
           ))
```

<br>


Creating plot
```{r}
# plotting total activity against age, faceted by education, and colored by gender
ggplot(activity_df, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(se = FALSE, linewidth = 1) +
  facet_wrap(~ education) +
  labs(
    title = "Total Daily Activity vs. Age by Education and Gender",
    subtitle = "Each point is one participant. Lines show smoothed trends.",
    x = "Age",
    y = "Total Daily Activity (Sum of MIMS)"
  )
```
<br>

Across all education levels, we see a general decline in total activity as participants get older. In the High School Equivalent group, we know there are significantly more males than females, therefore it appears that the females have a higher average activity level than males especially at a younger age. In the Less than High School group, the trend lines for both sexes are very close together and follow an almost identical pattern with a steep decline as age increases. Among those with a higher education, females have a consistently higher level of activity than males across almost the entire age range.

<br>
<br>

**Looking at 24-hour activity**

```{r}
# calculate average activity for each hour
hourly_activity = mims_df |> 
  mutate(
    minute = as.numeric(minute),
    hour_of_day = floor((minute - 1) / 60)
  ) |> 
  group_by(seqn, education, sex, hour_of_day) |> 
  summarise(
    mean_hourly_activity = mean(mims),
    .groups = 'drop'
  ) |>
  mutate(education = as.numeric(as.character(education)),
         education =
           case_match(
             education,
             1 ~ "Less than High School",
             2 ~ "High School Equivalent",
             3 ~ "More than High School"
           ))
  
```


Creating plot
```{r}
ggplot(hourly_activity, aes(x = hour_of_day, y = mean_hourly_activity, fill = sex)) +
  geom_col(position = "dodge") +
  facet_wrap(~ education, ncol = 2) +
  scale_x_continuous(
    name = "Time of Day (Hour)",
    breaks = c(0, 2, 4, 6, 8, 10, 12, 14, 16, 18, 20, 22)
  ) +
  
  labs(
    title = "Average Hourly Activity by Education Level and Gender",
    subtitle = "Activity has been averaged for each hour of the day.",
    y = "Mean Hourly MIMS Activity"
  )
```
<br>

The hourly activity in the "High School Equivalent" and "Less than High School" groups follow a similar pattern. Activity beings to rise around 6 or 7 AM and stays consistently high throughout the day with a gradual decline in the late evening. Within these two groups, there does not seem to be a significantly different pattern between males and females. In the "More than High School" group, females show a high peak of activity around 8 AM, while males show a high peak of activity in the late evening. 

















