---
title: "p8105_hw3_ab6168"
author: "Angelica Bailey"
date: "2025-10-06"
output: github_document
---

## Problem 1

Loading instacart dataset

```{r}
library(tidyverse)
library(p8105.datasets)
data("instacart")
```

Viewing dataset

```{r}
view(instacart)
```

**Description**:

The `instacart` data set contains information on grocery orders placed in 2017 by 131,209 unique users and includes 1,384,617 observations. There are a total of 15 variables describing the details of the order, product, and customer behavior. Each row in the data set is a product from an order. Key variables include:

-   `order_id` – Unique identifier for each order

-   `user_id` - Unique customer identifier

-   `product_id` and `product_name` - Identify the specific item purchased

-   `add_to_cart_order` - The sequence in which each product was added to the cart

-   `reordered` - Indicates whether product had been purchased by the user before

-   `order_dow` and `order_hour_of_day` - Represent when the order was placed

-   `aisle` and `department` - Describe the product's category

A few observations are that
#orders purchased on day 0
```{r}
instacart |> filter(
  order_dow == 0
)
```

